\begin{longtable}{p{16em}p{16em}}
    \textsc{Item}  & {\textsc{Date}} \\
    The original six points: & {{Announced by the Shah, January 9, 1963, and endorsed by national referendum, January 26, 1963}} \\
    1. Land reform (detailed in the 1962 land reform law) &  \\
    2. Nationalization of forests &  \\
    3. Sale of state-owned enterprises to the public &  \\
    4. Workers’ profit-sharing in 20 percent of net corporate earnings &  \\
    5. Voting and political rights for women &  \\
    6. Formation of the Literacy Corps &  \\
    Additional three points, 1964-65 &  \\
    7. Formation of the Health Corps & January 21, 1964 \\
    8. Formation of the Reconstruction and Development Corps & September 23, 1964 \\
    9. Establishment of Houses of Equity & October 13, 1965 \\
    Additional three points, 1967 & October 06, 1967 \\
    10. Program for nationalization of water resources &  \\
    11. Program for urban and rural reconstruction &  \\
    12. Administrative revolution (modernization, decentralization) &  \\
    Additional five points, 1975 &  \\
    13. Employee and public ownership extension scheme (up to 99 percent in state-owned enterprises and 49 percent in private firms) & September 09, 1975 \\
    14. Price stabilization and campaign against profiteering & September 09, 1975 \\
    15. Free education and a daily free meal for all children from kindergarten to eighth grade & \multicolumn{1}{p{11.715em}}{mid-December 1975} \\
    16. Free nutrition for infants up to the age of two & {late December 1975} \\
    17. Nation-wide social security (to be extended to rural population) & {late December 1975} \\
    18. Fight against land \& housing speculation &  \\
    19. Fight against corruption &  \\
    \end{longtable}%


TAKEN FROM: Iran Under the Pahlavis, George Lenczowski, ed. Hoover Institution Press: 1978. Reprinted by permission. 

SOURCES: Area Handbook for Iran, American University, Foreign Area Studies, 1971; Mohammad Reza Shah Pahlavi, The White Revolution, Tehran: 1967; Kayhan International; Quarterly Economic Review. 

